---
title: 'Introduction to Statistics: Final project'
author: "ARV"
date: "2022-11-15"
output:
  html_document:
    df_print: paged
---

# Spatial Autocorrelation in eletion data: Election for DK Parliamemt 2022

## Purpose
Investigate whether 1) there are significant differences between the average share of votes for the party the Social Democrats in the two voting regions Northern Jutland and the Capital Region, and 2) whether there is a significant spatial clustering of similar values.

```{r message=FALSE, warning=FALSE,results = 'hide'}
install.packages("tidyverse")
install.packages("sf")
install.packages("sp")
install.packages("spdep")
install.packages("rgdal")
install.packages("rgeos")
install.packages("tmap")
install.packages("tmaptools")
install.packages("spgwr")
install.packages("grid")
install.packages("gridExtra")
```
```{r message=FALSE, warning=FALSE, results = 'hide'}
library(tidyverse)
library(sf)
library(ggplot2)
library(sp)
library(spdep)
library(rgdal)
library(rgeos)
library(tmap)
library(tmaptools)
library(spgwr)
library(grid)
library(gridExtra)
library(pwr)

# Change the presentation of decimal numbers to 4 and avoid scientific notation
options(prompt="R> ", digits=4, scipen=999)
options(stringsAsFactors = FALSE)
```

### 1. Load and prepare data
```{r}
voting_data = st_read("data/voting.gpkg")
```
```{r}
head(voting_data, 10)
```
```{r}
summary(voting_data$A_pct)
```

**Plotting the distribution of share of votes for A shows that the values are approximately normally distributed**
```{r}
hist(voting_data$A_pct, 
     main = "Histogram of share of votes for the Social Democrats",
     xlab="Share of votes",
     ylab="Number of voting districts", 
     col="Purple", 
     border="Pink")
```
**Plot overview of the data**
```{r}
plot(voting_data["A_pct"], key.pos = 4, axes = FALSE, key.width = lcm(1.3), key.length = 1.,
     breaks = "jenks", lwd = 0.001, border = 'grey',main="Percent votes for the Social Democrats") 
```

```{r}
# Create new dataframe with a subset of the columns
voting_sel <- select(voting_data,c(Valgsted.navn, Landsdels.navn, A_pct, None_pct, geom))
```

```{r}
# Get pct votes for A in regions 'Hovedstaden' and 'Midtjylland-Nordjylland' as vectors
capital <- voting_sel[voting_sel$Landsdels.navn =="Hovedstaden",]
northern_jutland <- voting_sel[voting_sel$Landsdels.navn =="Midtjylland-Nordjylland",]

votes_c <- c(capital$A_pct)
votes_j <- c(northern_jutland$A_pct)
```

**Inspect subsets of data**
```{r}
hist(capital$A_pct,
     main = "Share of votes for the Social Democrats in the Capital Region",
     xlab="Share of votes",
     ylab="Number of voting districts", 
     col="Red", 
     border="Pink")
```
```{r}
summary(capital$A_pct)
```
```{r}
hist(northern_jutland$A_pct,
     main = "Share of votes for the Social Democrats in Northern Jutland",
     xlab="Share of votes",
     ylab="Number of voting districts", 
     col="Blue", 
     border="Pink")
```

```{r}
summary(northern_jutland$A_pct)
```


**Plot subsets**
```{r}
plot(capital["A_pct"], key.pos = 4, axes = FALSE, key.width = lcm(1.3), key.length = 1.,
     breaks = "jenks", lwd = 0.1, border = 'grey',main="Percent votes for the Social Democrats in the capital region") 
```

```{r}
plot(northern_jutland["A_pct"], key.pos = 4, axes = FALSE, key.width = lcm(1.3), key.length = 1.,
     breaks = "jenks", lwd = 0.1, border = 'grey',main="Percent votes for the Social Democrats in Northern Jutland") 
```

### 2. Analyse difference in average share of votes for the Social Democrats in Northen Jutland and the Capital Region.

To test whether the average share of votes for the Social Democrats in the two areas are significantly different (with alpha=0.05), a two-tailed test using the t-statistic for two independent samples/groups is performed.

Ho: The population means (average share of votes for A) are the same in the two regions
H1: The population means are different

```{r}
mean_c <- mean(votes_c)
mean_j <- mean(votes_j)

n1 <- length(votes_c)
n2 <- length(votes_j)

df <- n1+n2-2

s1 <- var(votes_c)
s2 <- var(votes_j)
```

To find t_critical, the qt()-function is used
```{r}
t_critical <- qt(0.05/2, df=df, lower.tail = TRUE)
t_critical
```
```{r}
t_critical_2 <- qt(0.01/2, df=df, lower.tail = TRUE)
t_critical_2
```
```

t_obtained can be calculated as t_obtained = (mean1 - mean2) / Sx1-x2
```{r}
Sx1_x2 <- sqrt((s1/n1)+(s2/n2))
t_obtained <- (mean_c - mean_j) / Sx1_x2
t_obtained
```

Since t_obtained exceeds the value for t_critical, the null hypothesis is rejected.

As a last step in this section, the confidence intervals are computed:
```{r}
CI_95_1 = (mean_c-mean_j) + t_critical * Sx1_x2
CI_95_2 = (mean_c-mean_j) - t_critical * Sx1_x2
CI_95_1
CI_95_2
```

*We can thus with 95% confidence say that the average share of votes for the Social Democrats is between -4.8 and -2.6 lower in the Captial region than in Northern Jutland.*

Power
```{r}
d <- (mean_c - mean_j / sd(voting_data$A_pct))
pow <- pwr.t2n.test(n1=261, n2=496, d=d, sig.level = 0.01,power=NULL,alternative="two.sided")
pow
```

### 3. Analyse spatial autocorrelation using [Global Moran and Moran's I](https://geographicdata.science/book/notebooks/07_local_autocorrelation.html)

The concept of 'spatial autocorrelation' is concerned with to extent to which similar values are located together - i.e. if there are spatial patterns in the distribution of values.
It can be compared to looking for correlation between two variables, but in this case one of the variables is the *location* of the objects being examined.
A common metric for investigating spatial autocorrelation is Moran's I.

The degree of spatial autocorrelation can both be computed at a global and at a local level. The global measure looks at whether similar values generally tends to be clustered together. The local measure detects local patterns in the distribution of values to answer the question of where the clusters are located.

Both global and local measures of spatial autocorrelation requires input data on which values are located next to each other ('neighbors') as well as a spatial weights matrix, which is used to provide the local average of values surrounding each value.

**Define queen neighbors**
```{r}
#voting_data <- tibble::rowid_to_column(voting_data, "ID") # unique ID required

stopifnot(length(unique(c(voting_data$ID))) == dim(voting_data)[1])

q_neigh <- poly2nb(voting_data, voting_data$ID)
q_neigh
```
```{r}
coords <- st_coordinates(st_centroid(st_geometry(voting_data))) # coordinates for plotting

plot(voting_data$geom, border = 'lightblue')
plot(q_neigh, coords, add=TRUE, col='purple', main='Queen Neighbors')
```
#### Global spatial autocorrelation

The global Moran's I value is between -1 and 1, where -1 indicates negative spatial autocorrelation (i.e. dissimilar values are next to each other), 0 a random pattern, and 1 positive spatial autocorrelation (similar values are next to each other.)

```{r}
# Create subsets without areas with no neighbors

voting_sub <- subset(voting_data, card(q_neigh) > 0)

neigh_sub <- subset(q_neigh, subset=card(q_neigh) > 0)
```

```{r}
# Define weights matrix
listw <- nb2listw(neighbours=neigh_sub)
listw
```
```{r}
# compute global moran for percent votes for A
globalMoran <- moran.test(voting_sub$A_pct, listw)
globalMoran
```
```{r}
globalMoran[["estimate"]][["Moran I statistic"]]
```

```{r}
globalMoran[["p.value"]]
```

*Since our Moran's I value is larger than zero and the p-value for our global Moran's I is very low, we can conclude that there is significant clustering of the examined value in our data.*

#### Local spatial autocorrelation

Local Moran's I are not confined to the -1 - 1 value range, but as with the global Moran's I, negative values indicate a lack of clustering while positive values indicate that similiar values are placed together. Just as with the global measure, we need to check whether the results are statistically significant.

```{r}
local_m <- localmoran(x = voting_sub$A_pct, listw = nb2listw(neigh_sub, style = "W"))
head(local_m)
```
```{r}
moran.map <- cbind(voting_sub, local_m)

tm_shape(moran.map) +
  tm_fill(col = "Ii",
          style = "quantile",
          title = "Local Moran's I statistic",
          palette = "PRGn",
          midpoint = NA) 
```
*From the plot we can conclude that some, but not all areas in Denmark, exhibits spatial autocorrelation when it comes to the share of votes for the Social Democrats.*


```{r}
quadrant <- vector(mode="numeric",length=nrow(local_m))

# centers the variable of interest around its mean
m.votes <- voting_sub$A_pct - mean(voting_sub$A_pct)     

# centers the local Moran's around the mean
m.local <- local_m[,1] - mean(local_m[,1])    

# significance threshold
sig <- 0.1 

# builds a data quadrant
quadrant[m.votes >0 & m.local>0] <- 4  # high-high
quadrant[m.votes <0 & m.local<0] <- 1  # low-low
quadrant[m.votes <0 & m.local>0] <- 2  # low-high
quadrant[m.votes >0 & m.local<0] <- 3  # high-low

# filter out insignificant values
quadrant[local_m[,5]>sig] <- 0 

# plot
brks <- c(0,1,2,3,4)
colors <- c("white","blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha=0.4),"red")
plot(voting_sub$geom,border="lightgray",col=colors[findInterval(quadrant,brks,all.inside=FALSE)])

box()
legend("topright", legend = c("not significant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")
```

*In the map above, dark red areas indicate areas with a high share of votes for the Social Democrats surrounded by other areas with a high share of votes for the party, while dark blue areas indicate a similar pattern but for low share of votes. Pale red indicate an area with a high share of votes surrounded by lower share of votes, while pale blue indicate low share of votes surrounded by higher shares of votes.*

